---
title: "STAT 847: Midterm Project"
subtitle: "Chris Binoi Verghese ID:21092999"
output:
  pdf_document: default
  html_document: default
urlcolor: blue
---



```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r, include = FALSE}
library(knitr)
library(plyr)
library(tidyverse)
library(ggplot2)
#opts_chunk$set(tidy.opts=list(width.cutoff=70),tidy=TRUE)
```




Q1. (20 points) Use the elements column of the **detailed dataset** to infer about the differences in blinking and jaw clenching between resting and non-resting states. Specifically, find differences in the distribution of the number of blinks and jaw clenches per minute during resting and active sessions.

```{r message=F, warning=F, fig.height=10, fig.width=10}
dat_all = read.csv("C:/Users/chris/Desktop/MDSAI/STAT847/Midterm/Mind Monitor detailed data 2024-01-21.csv")
dat_summary = ddply(dat_all, "sessionnum", summarise,
                    activity = activity[1],
                    session_time = session_time[1],
                    mean_alpha = mean(total_alpha, na.rm=TRUE),
                    mean_beta = mean(total_beta, na.rm=TRUE),
                    mean_gamma = mean(total_gamma, na.rm=TRUE),
                    mean_delta = mean(total_delta, na.rm=TRUE),
                    mean_theta = mean(total_theta, na.rm=TRUE),
                    var_alpha = var(total_alpha, na.rm=TRUE),
                    var_beta = var(total_beta, na.rm=TRUE),
                    var_gamma = var(total_gamma, na.rm=TRUE),
                    var_delta = var(total_delta, na.rm=TRUE),
                    var_theta = var(total_theta, na.rm=TRUE),
                    blinks_minute = Nblinks[1]/session_time[1]*60,
                    jaws_minute = Njaw[1]/session_time[1]*60,
                    mean_pos_xy = mean(sqrt(Accelerometer_X^2 + Accelerometer_Y^2),
                                       na.rm=TRUE),
                    mad_accel = mean(abs(Accelerometer_X^2 + Accelerometer_Y^2 + Accelerometer_Z^2 - 1), 
                                       na.rm=TRUE),
                    rmse_gyro = mean(sqrt(Gyro_X^2 + Gyro_Y^2 + Gyro_Z^2), na.rm=TRUE)
)

dat_rest = dat_summary %>% filter(activity == 'resting')
dat_active = dat_summary %>% filter(activity != 'resting')
summary(dat_active[c("blinks_minute","jaws_minute")])
summary(dat_rest[c("blinks_minute","jaws_minute")])


```

The summary analysis of the number of blinks and jaw clenches per minute reveals interesting patterns between resting and active sessions.

For the number of blinks per minute, the summary statistics show that during active sessions, there is a significantly higher frequency of blinks. This is evident from the smaller minimum, maximum, median, and the extremely small 3rd quantile value during resting sessions compared to the active sessions. The resting sessions exhibit a much narrower range of blink frequencies, suggesting a more consistent pattern during activity.

In contrast, the number of jaw clenches per minute follows a different trend. While the minimum and maximum values are higher during resting sessions, the overall median and 3rd quantile values are higher during active sessions. This indicates that the distribution of jaw clenches per minute is skewed by an extremely high maximum value during one resting session, while the distribution during active sessions is more stable.

To visualize these distributions, Kernel Density Estimation (KDE) is used, providing a smooth estimate of the probability density function of the data. This non-parametric method helps in understanding the distribution patterns more clearly, especially in cases where standard summary statistics might not capture the nuances of the data distribution.

```{r message=F, warning=F, fig.height=10, fig.width=10}
# Set up a 2x2 layout for the plots
par(mfrow = c(2, 1))
plot(density(dat_rest$blinks_minute),
     main = "Kernel Density Estimation for blinks per minute at rest", 
     col = "red")
plot(density(dat_active$blinks_minute), 
     main = "Kernel Density Estimation for blinks per minute when active", 
     col = "blue")
plot(density(dat_rest$jaws_minute), 
     main = "Kernel Density Estimation for jaw clenches per minute at rest", 
     col = "red")
plot(density(dat_active$jaws_minute), 
     main = "Kernel Density Estimation for jaw clenches per minute when active", 
     col = "blue")
```


The Kernel Density Estimation (KDE) graphs provide insights into the frequency distribution of jaw clenches and blinks per minute during resting and active sessions. In resting sessions, both the number of jaw clenches and blinks per minute exhibit a wider range of values, with variations between very high and low frequencies.

In contrast, during active sessions, the KDE for blinks is more bell-shaped, indicating a more concentrated distribution around a specific frequency, while the distribution for jaw clenches during an active session resembles the distribution of blinks, suggesting a similar behavior pattern.

This analysis suggests that people tend to blink more frequently during active sessions compared to resting sessions. Additionally, the behavior of jaw clenching, though more defined in rest sessions, shows similarities between resting and active states.

\newpage

Q2.(15 points) Use the data in the **summary dataset** to infer about the differences in brainwave activity between resting and non-resting states. Specifically, conduct a t-test of each of the ten brainwave variables to find any statistically significant differences. To get the p-value from a test, use `t.test(...)$p.value`.
```{r}
#Read Summary dataset
dat_sum = read.csv("C:/Users/chris/Desktop/MDSAI/STAT847/Midterm/Mind Monitor session summaries 2024-01-21.csv")
dat_rest = dat_sum %>% filter(activity == 'resting')
dat_active = dat_sum %>% filter(activity != 'resting')
# List of features for which you want to perform t-tests
features <- c("mean_alpha", "mean_beta", "mean_gamma", "mean_delta", "mean_theta",
              "var_alpha", "var_beta", "var_gamma", "var_delta", "var_theta")

# Create an empty data frame to store results
p_values_df <- data.frame(Feature = character(), P_Value = numeric(), stringsAsFactors = FALSE)

# Loop through each feature and perform t-test
for (feature in features) {
  p_value <- t.test(dat_rest[[feature]], dat_active[[feature]])$p.value
  p_values_df <- rbind(p_values_df, data.frame(Feature = feature, P_Value = p_value))
}

knitr::kable(p_values_df, col.names = c("brainwave variable", "p-value"), 
             caption = "p-Values for brain wave means and variances")
```


In our attempt to assess the similarity of data distribution for various brainwave variables (including "mean_alpha," "mean_beta," "mean_gamma," "mean_delta," "mean_theta," "var_alpha," "var_beta," "var_gamma," "var_delta," and "var_theta") derived from the summary dataset, which represents means and variances of different brainwaves over each session, we employed a Welch two-sample t-test. This test aimed to determine whether there are any statistically significant differences between the brainwave values during resting and active sessions.

Upon conducting the t-test, we discovered that the mean value of the total gamma waves is the only brainwave variable exhibiting a statistically significant difference. The calculated p-value for the t-test on the mean of gamma waves during active and resting sessions is 0.00039, an exceptionally low value. Generally, a p-value of 0.05 is considered a suitable threshold to assess whether a null hypothesis of similar means can be rejected. In comparison, the next closest p-value is for the mean of alpha waves, which is 0.0687—slightly lower than the other variables and in proximity to the 0.05 threshold.

Therefore, the brainwave variable that distinctly demonstrates statistically significant differences between resting and non-resting states is **mean gamma**.

\newpage
Q3. (10 points) Are there any notable differences between the first 20% and the last 20% (by time_in) of active sessions? Look in the **detailed dataset** and isolate the first active session to find out. (The answer may be 'no'. In which case, mention some similarities). Compare at least three variables.
```{r}
dat_active = dat_all %>% filter(activity != 'resting')
min_sessionnum <- min(dat_active$sessionnum)
# Subset the data frame for rows where sessionnum is the minimum and activity is resting
isolated_values <- dat_active[dat_active$sessionnum == min_sessionnum, ]
max_time = max(isolated_values$time_in)
first_20 <- isolated_values[isolated_values$time_in < max_time*0.2, ]
last_20 <- isolated_values[isolated_values$time_in > max_time*0.8, ]
p_values <- c(
  t.test(first_20$total_alpha, last_20$total_alpha)$p.value,
  t.test(first_20$total_beta, last_20$total_beta)$p.value,
  t.test(first_20$total_gamma, last_20$total_gamma)$p.value,
  t.test(first_20$total_delta, last_20$total_delta)$p.value,
  t.test(first_20$total_theta, last_20$total_theta)$p.value
)

# Create a data frame to display the p-values
results_df <- data.frame(
  Brainwave_Band = c("total_alpha", "total_beta", "total_gamma", "total_delta", "total_theta"),
  P_Value = p_values
)
knitr::kable(results_df, caption = "p-Values for data")

```


The p-values obtained from a Welch two-sample t-test provide insights into the statistically significant differences between the first 20% and last 20% of the first active session, considering the variables 'total_alpha', 'total_beta', 'total_gamma', 'total_delta', and 'total_theta'. Notably, there is a discernible difference between the datasets for all brainwaves, except for total alpha brain waves.

The standard p-value threshold considered here is 0.05. However, most of the brainwaves do not approach this value, except for total alpha and theta brainwaves. Total theta waves have a p-value of 0.0114, while total alpha waves exhibit a reasonably higher p-value. To visualize these changes more clearly, histograms of these values can be examined. 
```{r}
column_names <- c('total_alpha', 'total_beta', 'total_gamma', 'total_delta', 'total_theta')

# Set up the layout for the plot
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

# Loop through each column and create overlaid histograms
for (col in column_names) {
  hist(first_20[[col]], col = rgb(0.2, 0.8, 0.2, alpha = 0.5), 
       main = paste("Histogram of", col), xlab = col, 
       ylim = c(0, max(hist(first_20[[col]], plot = FALSE)$counts,
                       hist(last_20[[col]], plot = FALSE)$counts)))
  hist(last_20[[col]], col = rgb(0.8, 0.2, 0.2, alpha = 0.5), add = TRUE)
  legend("topright", legend = c("first_20", "last_20"), 
         fill = c(rgb(0.2, 0.8, 0.2, alpha = 0.5), rgb(0.8, 0.2, 0.2, alpha = 0.5)))
}
```


The plotted histograms reveal that the distribution of total alpha waves exhibits an extremely good overlap, indicating very similar means. However, this is not the case for the rest of the total values, except for total theta waves, which approach a similar mean. The distribution of total beta waves for the last 20% is slightly to the left and has a lower frequency compared to the first 20%. In contrast, the first 20% has a distribution with very similar but much smaller frequency peaks compared to the last 20%. Total delta waves show almost two max frequency peaks, suggesting a very different distribution, while total theta has a distribution where the mean of the last 20% is slightly to the right of the first 20%.

In summary, the most statistically similar distributions are observed in the first and last 20% of **total alpha waves** for the first active session.

\newpage
Q4. (10 points) Are there any notable differences between the first 20% and the last 20% (by time_in) of resting sessions? Look in the **detailed dataset** and isolate the first resting session to find out. (The answer may be 'no'. In which case, mention some similarities). Compare at least three variables.
```{r}
dat_rest = dat_all %>% filter(activity == 'resting')
min_sessionnum <- min(dat_rest$sessionnum)
# Subset the data frame for rows where sessionnum is the minimum and activity is resting
isolated_values <- dat_rest[dat_rest$sessionnum == min_sessionnum, ]
max_time = max(isolated_values$time_in)
first_20 <- isolated_values[isolated_values$time_in < max_time*0.2, ]
last_20 <- isolated_values[isolated_values$time_in > max_time*0.8, ]
p_values <- c(
  t.test(first_20$total_alpha, last_20$total_alpha)$p.value,
  t.test(first_20$total_beta, last_20$total_beta)$p.value,
  t.test(first_20$total_gamma, last_20$total_gamma)$p.value,
  t.test(first_20$total_delta, last_20$total_delta)$p.value,
  t.test(first_20$total_theta, last_20$total_theta)$p.value
)

# Create a data frame to display the p-values
results_df <- data.frame(
  Brainwave_Band = c("total_alpha", "total_beta", "total_gamma", "total_delta", "total_theta"),
  P_Value = p_values
)
knitr::kable(results_df, caption = "p-Values for data")

```

In resting sessions, the Welch two-sample t-test reveals significant differences in the distributions of "total_alpha," "total_beta," "total_gamma," "total_delta," and "total_theta" values. None of the p-values approach the set threshold of 0.05, with the closest being for total delta waves at 0.00156. Histograms visually illustrate these distinctions in wave distributions.
```{r}
column_names <- c('total_alpha', 'total_beta', 'total_gamma', 'total_delta', 'total_theta')

# Set up the layout for the plot
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

# Loop through each column and create overlaid histograms
for (col in column_names) {
  hist(first_20[[col]], col = rgb(0.2, 0.8, 0.2, alpha = 0.5), 
       main = paste("Histogram of", col), xlab = col, 
       ylim = c(0, max(hist(first_20[[col]], plot = FALSE)$counts, 
                       hist(last_20[[col]], plot = FALSE)$counts)))
  hist(last_20[[col]], col = rgb(0.8, 0.2, 0.2, alpha = 0.5), add = TRUE)
  legend("topright", legend = c("first_20", "last_20"), 
         fill = c(rgb(0.2, 0.8, 0.2, alpha = 0.5), rgb(0.8, 0.2, 0.2, alpha = 0.5)))
}

```


The distributions of total alpha and beta waves show a slight leftward shift in the last 20% of the session compared to the first 20%. Similarly, total gamma waves exhibit a more spread-out distribution in the first 20%, concentrating between 0 and -0.75 in the last 20%. Total delta waves display a rightward shift in frequency distribution in the last 20%, while total theta waves reach a maximum frequency at a lower total gamma wave value in the last 20% compared to the first 20%.

In summary, the total brainwave distributions in the first and last 20% of the resting session exhibit various shifts and differences, and there are no statistically significant similarities between the two segments.
\newpage

Q5. (5 points) Use the **detailed dataset** Make a variable in the summary data that gets the Pearson correlation between Alpha_TP9 and Alpha_TP10 during a session. (That is, the alpha waves above the left and right ear, respectively). Compare the TP9-TP10 correlations during in resting and non-resting activities.
```{r}
cor_summary <- dat_all %>%
  na.omit() %>%  # Remove rows with NA values
  group_by(sessionnum) %>%
  summarise(
    activity = first(activity),
    cor_value = {
      # Check for zero standard deviations before calculating correlation
      if (sd(Alpha_TP9) != 0 & sd(Alpha_TP10) != 0) {
        cor(Alpha_TP9, Alpha_TP10)
      } else {
        NA  # Set correlation to NA if standard deviation is zero
      }
    }
  )
knitr::kable(cor_summary, 
             col.names = c("session", "activity", "correlation"),
             caption = "Data Summary with Correlation of Alpha")
```


The correlation between alpha brainwaves over the left and right ears provides insights into their directional relationship at any given time. During resting sessions, there is a notable increase in correlation compared to active sessions. It is important to note that during active sessions, the correlation does not follow a clear trend, as the nature of the activities during these sessions impacts the direction of the alpha waves.

Notably, the presence of NA values in the correlation during active sessions, particularly when the user was gaming, suggests instances where the sensor was unable to capture the alpha brainwaves around the ears accurately. This indicates potential challenges in reading alpha brainwave data during certain activities.

\newpage
Q6. (5 points) The accelerometer always shows some acceleration because of gravity. (When completely upright and still, it shows accelerometer Z = 1, X = 0, Y = 0) You can use this to determine the relative position of the head, as well as how much the head moves. Use the acceleration data in the **detailed dataset** to infer about the differences in both the position of the head, and its movements, between resting and non-resting states.

e.g., you final answer should be something like "Head position was more _______ when resting than when active, and we know this because the accelerometer ________ instead of __________", with code and supporting summary information.
```{r}
dat_accelerometer <- dat_all %>%
  na.omit() %>%
  group_by(sessionnum) %>% 
  summarise(
    activity = first(activity),
    mean_X = mean(Accelerometer_X),
    mean_Y = mean(Accelerometer_Y),
    mean_Z = mean(Accelerometer_Z)
  )
knitr::kable(dat_accelerometer,
             caption = "Accelerometer Information for every session")
```

Head position was more towards the bottom right while being lower when resting than when active, and we know this because the accelerometer readings had a mean X value of about negative -0.6 while the mean z value was lower and the mean y value is always positive and more than 0.1 instead of in the case of being active where the z-value tends to be about 0.9 most of the time while the x-value tends to have a minimum of -0.145 which is more than any other x-value when active while the y value has a maximum right tilt of 0.068 while active.

From these observations, we can infer that the user's resting position involves tilting the head to the right and lowering it. This is supported by the lower Z values and more negative X values and more positive y values recorded by the accelerometer during rest sessions while compared to active sessions.

\newpage
Q7. (10 points) In the **detailed dataset**, plot the smoothed brainwaves for all five bands (alpha, beta, gamma, delta, and theta) session number 11 (resting, 3222 seconds long). Use any smoothing technique we discussed in class. Make a guess at when the patient fell asleep.
```{r}
# Subset data for session number 11
dat_11 <- dat_all %>% filter(sessionnum == 11  & !is.na(Delta_TP9))

## Smooth over time
ss_alpha = smooth.spline(dat_11$time_in, dat_11$total_alpha, spar=0.6)
ss_beta = smooth.spline(dat_11$time_in, dat_11$total_beta, spar=0.6)
ss_gamma = smooth.spline(dat_11$time_in, dat_11$total_gamma, spar=0.6)
ss_delta = smooth.spline(dat_11$time_in, dat_11$total_delta, spar=0.6)
ss_theta = smooth.spline(dat_11$time_in, dat_11$total_theta, spar=0.6)
plot(ss_alpha, type='n', col="White", lwd=3, ylim=c(-2,3),
     main="Brainwaves Over Time", xlab = "Time", ylab = "Smoothed Brainwave Value")
lines(ss_alpha, type='l', col="Red", lwd=3)
lines(ss_beta, type='l', col="Blue", lwd=3)
lines(ss_gamma, type='l', col="Black", lwd=3)
lines(ss_delta, type='l', col="Darkgreen", lwd=3)
lines(ss_theta, type='l', col="Grey", lwd=3)
# Add legend
legend("topright", legend = c("Alpha", "Beta", "Gamma", "Delta", "Theta"),
       col = c("Red", "Blue", "Black", "Darkgreen", "Grey"), lty = 1, lwd = 2.5, cex = 0.8)
```

Here is the smoothed representation of the patient's brainwaves during session 11. A notable observation is a significant dip in all brainwave frequencies, except for gamma waves, occurring at approximately **750 seconds**. This distinctive pattern strongly suggests that the patient entered a sleep state around this time. Another supporting factor for this assumption is the remarkably stable brainwave activity observed shortly after the onset of sleep, in contrast to the fluctuations observed before falling asleep.


\newpage
Q8.(10 points)  Note that total_alpha and its similar functions have a flaw - measurements that were missing because of a bad connection are marked as zero, while true values are rarely zero. Find another measure using the **detailed dataset**, mean_alpha, that takes the average of **all the non-zero alpha measurements**.  Take the mean of this new measurement for each session. For example, the average of mean_alpha for all of session 1 as a single value, all of session 2 as a single value, and so on.
```{r}


detailed_data <- dat_all %>%
  mutate(
    counter = (Alpha_TP9 != 0) + (Alpha_AF7 != 0) + (Alpha_AF8 != 0) + (Alpha_TP10 != 0),
    mean_alpha = (Alpha_TP9 + Alpha_AF7 + Alpha_AF8 + Alpha_TP10) / counter
  )

results <- detailed_data %>%
  group_by(sessionnum) %>%
  summarise(
    activity  = first(activity),
    average_alpha = mean(mean_alpha, na.rm = TRUE)
  )

knitr::kable(results, 
             col.names = c("session", "activity", "Mean Alpha"), 
             caption = "Data Summary for total means")
```



\newpage
Q9.(15 points) Find two interesting patterns in the data. These could be almost anything. You can use **either the detailed or the summary datasets**.

The first interesting pattern of data involved the frequency of number of blinks and number of jaw clenches in each session in relation to its activity and each other.

```{r message=F, warning=F, fig.height=3, fig.width=10}

# Get unique sessionnum values
unique_sessionnums <- unique(dat_all$sessionnum)

# Loop through each sessionnum
for (k in unique_sessionnums) {
  session_data <- dat_all %>% filter(sessionnum == k)
  
  # Create a density plot for N-blinks and N-jaws
  plot <- ggplot(session_data, aes(x = `Nblinks`, fill = "Nblinks")) +
    geom_density(alpha = 0.5, color = "blue") +
    geom_density(aes(x = `Njaw`, fill = "Njaw"), alpha = 0.5, color = "green") +
    labs(title = paste("Density Plot for session: ", 
                       k, "  Activity:", unique(session_data$activity)),
         x = "Count",
         y = "Density",
         fill = "Variable") +
    scale_fill_manual(values = c("blue", "green")) +
    theme_minimal()
  
  # Print each plot individually
  print(plot)
}

```


In general, when the patient is engaged in active tasks, there is a tendency for a higher frequency of blinks compared to jaw clenches. Conversely, during resting sessions, there is a prevalence of jaw clenches compared to blinks. However, the nature of the activity significantly influences this pattern.

During reading, despite the patient being actively engaged, the trend deviates. Jaw clenches surpass blinks in number, and their distribution is more evenly spread rather than concentrated around extremes observed in typical resting and active sessions. This suggests a unique pattern during reading, where the patient exhibits a moderate and evenly distributed jaw-clenching activity.

In session 4, characterized by gaming, an extreme occurrence of both jaw clenches and blinks is observed. Interestingly, instances of no blinking at all are present, contributing to the distinctive trend during gaming sessions.

Watching TV generally follows the pattern of regular active sessions, with a higher number of blinks compared to jaw clenches. However, in the session immediately following TV watching and preceding resting, the trend aligns more with active sessions. Here, the number of jaw clenches significantly exceeds the number of blinks, implying that the patient might be reflecting on or processing the content viewed, rather than engaging in a specific activity.

During session number 18, where the patient is listening to music, a more relaxed state is apparent. Fewer jaw clenches are observed, and there is a higher frequency of blinks spread over a broader range. This suggests a calmer and less physically active state during music listening compared to other active sessions.
```{r message=F, warning=F, fig.height=3, fig.width=10}
dat_EEG <- dat_all %>%
  na.omit() %>%
  group_by(sessionnum) %>%
  summarise(
    sessionnum = first(sessionnum),
    activity = first(activity),
    mean_eeg = mean(AUX_RIGHT),
    std_eeg = sd(AUX_RIGHT)
  )
ggplot(dat_EEG, aes(x = sessionnum, y = mean_eeg)) +
  geom_line(color = 'green') +
  geom_point(aes(color = ifelse(activity == 'resting', 'red', 'blue'))) +
  labs(title = "Mean Raw EEG Over Sessions with Red Resting values",
       x = "Session Number",
       y = "Mean EEG") +
  scale_color_identity() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(dat_EEG, aes(x = sessionnum, y = std_eeg)) +
  geom_line(color = 'green') +
  geom_point(aes(color = ifelse(activity == 'resting', 'red', 'blue'))) +
  labs(title = "Standard Deviation of Raw EEG Over Sessions with Red Resting values",
       x = "Session Number",
       y = "Standard Deviation of EEG") +
  scale_color_identity() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


The first graph depicts the mean Raw EEG sensor values in microvolts from AUX_RIGHT in the detailed dataset. The red points correspond to mean EEG values during resting, while the blue points represent active periods.

It is noteworthy that the mean EEG consistently decreases during each resting session across all sessions. Session 9, during which the patient watched TV, exhibits a noticeable dip in mean EEG. Conversely, in session 13, characterized as an active session, there is a spike in EEG before transitioning to resting. Notably, two subsequent resting sessions feature elevated EEG levels, standing out from the established pattern. Conversely, session 13, following a meeting, emerges as an outlier, deviating from the usual trend observed in prior sessions.

The second plot illustrates the standard deviation of Raw EEG sensor values in microvolts from AUX_RIGHT in the detailed dataset. Red points represent mean EEG values during resting, while blue points signify active periods.

In this plot, the standard deviation of Raw EEG from the mean appears relatively uniform, with exceptions observed at session 9 and session 13. Session 9's variation can be attributed to the patient watching TV, where EEG dynamics are likely influenced by the content viewed. Session 13 stands out as an anomalous session, as noted in the previous graph. This active session, following a meeting, displays distinctive EEG variability, suggesting a different activity level compared to previous sessions.
\newpage